Calibrating MEM seeding heuristics

### 2018-08-10 MEM.tex
Fixed the theory (and checked exactness through implementation).

With this notation, the probability that at least one mask survives a
sequence of $i$ error-free nucleotides is thus $1-\xi_i^N$, and the
probability that there remains a mask at the $i-1$-th but not at the 
$i$-th error-free nucleotide is $\xi_i^N - \xi_{i-1}^N$. From this we
conclude that the weighted generating function of the segments terminated
by $\uparrow_{/i}$ following a segment terminated by $\downarrow_{/m}$ is
\begin{equation}
\label{eq:B}
B_i(z) = \Big( \xi_{i-1}^N-\xi_i^N \Big) (qz)^i.
\end{equation}

@@

The probability of condition $ii)$ is
\begin{equation*}
\xi_i^m \Big(1 - \eta_i^{N-m} \Big),
\end{equation*}
but we need to break up this term among the terminators $\downarrow_{/n}$,
($0 \leq n \leq N$). For this, we need to keep track of the number of soft
masks that run to the end of the segment, including the terminator. The
probability that there are $r \geq 1$ such soft masks is
\begin{equation*}
{N-m \choose r} (1 - \eta_i)^r \eta_i^{N-m-r}.
\end{equation*}

Each of the terminator, so the total number of matches is $r$ plus the
number of sequences, among the $N-r$ that remain, that also match the
terminator. The probability that in total $n$ sequences match the
terminator is
\begin{eqnarray*}
&\;& \sum_{r\geq1} {N-m \choose r} (1 - \eta_i)^r
\eta_i^{N-m-r} {N-r \choose n-r} (\mu/3)^{n-r} (1-\mu/3)^{N-n} \\
&=& \eta_i^{N-m} (\mu/3)^n(1-\mu/3)^{N-n}
  \sum_{r\geq1} {N-m \choose r} {N-r \choose n-r}
  \left( \frac{1-\eta_i}{\eta_i \mu/3} \right)^r \\
&=& \omega_n \cdot \eta_i^{N-m} \cdot \zeta_{i,m,n},
\end{eqnarray*}
where
\begin{equation}
\label{eq:zeta}
\zeta_{i,m,n} = \sum_{r\geq1}
{N-m \choose r} {N-r \choose n-r}
\left( \frac{1-\eta_i}{\eta_i \mu/3} \right)^r \bigg/ {N \choose n}.
\end{equation}

%\begin{remark}
%It is possible to approximate $\zeta_{i,m,n}$ with a simpler expression as
%follows. We start with general expressions of the form
%\begin{equation*}
%\sum_{r\geq0} {N-m \choose r} {N-r \choose n-r} x^r.
%\end{equation*}
%
%Applying $m$ times the identity ${a \choose b} = {a-1 \choose b} + {a-1
%\choose b-1}$ we obtain
%\begin{equation*}
%{N-m \choose n-r} = \sum_{s=0}^m {m \choose s} {N-r-m \choose n-r-s}.
%\end{equation*}
%
%Combining the binomial terms, we can see that
%\begin{equation*}
%{N-m \choose r} {N-r-m \choose n-r-s} = {N-m \choose n-s} {n-s \choose r}.
%\end{equation*}
%
%This finally leads to the following expression featuring a sum of
%hypergeometric probabilities
%\begin{equation*}
%\sum_{s=0}^m {m \choose s} {N-m \choose n-s}  \sum_{r\geq0} {n-s
%\choose r} x^r = \sum_{s=0}^m {m \choose s} {N-m \choose n-s} (1+x)^{n-s}.
%\end{equation*}
%
%The hypergeometric distribution can be approximated by the binomial
%distribution using the following formula
%\begin{equation*}
%{m \choose s} {N-m \choose n-s} \Bigg/ {N \choose n} \approx
%{n \choose s} (m/N)^s (1-m/N)^{n-s}.
%\end{equation*}
%
%We can thus simplify the sum above using Newton's binomial theorem. We
%obtain
%\begin{equation*}
%{N \choose n} \big( 1+x \cdot (1-m/N) \big)^n.
%\end{equation*}
%
%From this we conclude
%\begin{equation*}
%\sum_{r=1}^{N-m} {N-m \choose r} {N-r \choose n-r} x^r  \approx
%{N \choose n} \Big[ \big( 1+x \cdot (1-m/N) \big)^n - 1 \Big].
%\end{equation*}
%
%Substituting the value of $x$, we finally obtain
%\begin{equation*}
%\zeta_{i,m,n} \approx \omega_n \xi_i^m \cdot \lambda_{i,m,n},
%\end{equation*}
%where
%\begin{equation}
%\lambda_{i,m,n} = \eta_i^{N-m} \left[ \left( 1+ \frac{1-\eta_i}{\eta_i \mu
%/ 3} \cdot \frac{N-m}{N} \right)^n - 1 \right].
%\end{equation}
%\end{remark}

The weighted generating function we are looking for is thus
\begin{equation}
\label{eq:A}
A_{m,n} =
\omega_n pz \sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i + \omega_n
pz \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \cdot (1- \eta_i^{N-m}
\zeta_{i,m,n}) \Big) (qz)^i.
\end{equation}

@@

The final expression for the transfer matrix $M(z)$ is
\begin{equation*}
\begin{blockarray}{cccccccc}
   & \dn{0} & \ldots & \dn{N} & \up{1} & \ldots & \up{\gamma-1} & \nd \\
\begin{block}{c[ccccccc]}
\dn{0} & A_{0,0}(z) & \ldots & A_{0,N}(z) & B_1(z) & \ldots &
    B_{\gamma-1}(z) & C_0(z) \\
\vdots & \vdots & \ddots & \vdots & \vdots & \ddots &
    \vdots & \vdots \\
\dn{N} & A_{N,0}(z) & \ldots & A_{N,N}(z) & B_1(z) & \ldots &
    B_{\gamma-1}(z) & C_N(z) \\
\up{1} & D_{1,0}(z) & \ldots & D_{1,N}(z) & 0 & \ldots & 0 & E_1(z) \\
\vdots & \vdots & \ddots & \vdots & \vdots & \ddots &
    \vdots & \vdots \\
\up{\gamma-1} & D_{\gamma-1,0}(z) & \ldots & D_{\gamma-1,N}(z) & 0 &
  \ldots & 0 & E_{\gamma-1}(z) \\
\nd & 0 & \ldots & 0 & 0 & \ldots & 0 & 0 \\
\end{block}
\end{blockarray}
\end{equation*}
where
\begin{gather}
\tag{\ref{eq:A}}
A_{m,n} =
\omega_n pz \sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i + \omega_n
pz \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \cdot (1- \eta_i^{N-m}
\zeta_{i,m,n}) \Big) (qz)^i \\
\tag{\ref{eq:B}}
B_i(z) = \Big( \xi_{i-1}^N-\xi_i^N \Big) (qz)^i \\
\tag{\ref{eq:C}}
C_m(z) =
\sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i +
  \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \Big) (qz)^i \\
\tag{\ref{eq:D}}
D_{j,m}(z) = \omega_m pz \sum_{i=0}^{\gamma-j-1} (qz)^i \\
\tag{\ref{eq:E}}
E_j(z) = \sum_{i=0}^{\gamma-j-1} (qz)^i
\end{gather}
and where
\begin{gather}
\tag{\ref{eq:omega}}
\omega_m = {N \choose m} \big(1 - \mu/3\big)^{N-m} \big(\mu/3\big)^m \\
\tag{\ref{eq:xi}}
\xi_i = 1-(1-\mu)^i \\
\tag{\ref{eq:eta}}
\eta_i = 1-(1-\mu)^i\mu/3 \\
\tag{\ref{eq:zeta}}
\zeta_{i,m,n} = \sum_{r\geq1}
{N-m \choose r} {N-r \choose n-r}
\left( \frac{1-\eta_i}{\eta_i \mu/3} \right)^r \bigg/ {N \choose n}.
\end{gather}

@@

>>>>

With this notation, the probability that at least one mask survives a
sequence of $i$ error-free nucleotides is thus $1-\xi_i^N$, and the
probability that there remains a mask at the $i-1$-th but not at the 
$i$-th error-free nucleotide is $\xi_i^N - \xi_{i-1}^N$. From this we
conclude that the weighted generating function of the segments terminated
by $\uparrow_{/i}$ following a segment terminated by $\downarrow_{/m}$ is
\begin{equation}
\label{eq:B}
B_i(z) = \Big( \xi_i^N-\xi_{i-1}^N \Big) (qz)^i.
\end{equation}

@@

The probability of condition $ii)$ is
\begin{equation*}
\xi_i^m \Big(1 - \eta_i^{N-m} \Big),
\end{equation*}
but we need to break up this term among the terminators $\downarrow_{/n}$,
($0 \leq n \leq N$). For this, we decompose the sum on the number of soft
masks that run to the end of the segment, including the terminator. The
probability that there are $r \geq 1$ such soft masks is
\begin{equation*}
{N-m \choose r} (1 - \eta_i)^r \eta_i^{N-m-r}.
\end{equation*}

Each of them matches the nucleotide at the end of the segment, so the
total number of matches is $r$ plus the number of sequences, among the
remaining $N-m-r$ soft masks and $m$ hard masks, that also match the
last nucleotide.

Let us start with the $N-m-r$ that failed somewhere in the segment. The
conditional probability that they match the last nucleotide given that
they fail somewhere in the segment is $\mu/3 \cdot \xi_i / \eta_i$. The
conditional probability that they do not match is $(1-\mu/3) / \eta_i$.
For the $m$ hard masks, the probability that they match the last
nucleotide given that they sail somewhere in the segment is simply
$\mu/3$, and the probability that they do not is $1-\mu/3$.

The probability that those $N-r$ duplicates contribute $n-r$ matches is
\begin{equation*}
\frac{(\mu/3)^{n-r}(1-\mu/3)^{N-n}}{\eta_i^{N-m-r}}
\psi_{i,m,n,r}\text{, where \;}
\psi_{i,m,n,r} = \sum_{q \geq 0}{m \choose q}{N-m-r \choose n-r-q}
\xi_i^{n-r-q}.
\end{equation*}

The probability that in
total $n$ sequences match the terminator is
\begin{eqnarray*}
&\;& \sum_{r\geq1} {N-m \choose r}
(1 - \eta_i)^r (\mu/3)^{n-r} (1-\mu/3)^{N-n} \psi_{i,m,n,r} \\
&=& (\mu/3)^n(1-\mu/3)^{N-n} \sum_{r\geq1} {N-m \choose r}
  (1 - \mu)^{ri} \psi_{i,m,n,r} \\
&=& \omega_n \cdot \zeta_{i,m,n},
\end{eqnarray*}
where
\begin{equation}
\label{eq:zeta}
\zeta_{i,m,n} = \sum_{r\geq1} {N-m \choose r}
(1-\mu)^{ri} \psi_{i,m,n,r} \bigg/ {N \choose n}.
\end{equation}

The weighted generating function we are looking for is thus
\begin{equation}
\label{eq:A}
A_{m,n} =
\omega_n pz \sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i + \omega_n
pz \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \cdot
(1- \zeta_{i,m,n}) \Big) (qz)^i.
\end{equation}

@@

The final expression for the transfer matrix $M(z)$ is
\begin{equation*}
\begin{blockarray}{cccccccc}
   & \dn{0} & \ldots & \dn{N} & \up{1} & \ldots & \up{\gamma-1} & \nd \\
\begin{block}{c[ccccccc]}
\dn{0} & A_{0,0}(z) & \ldots & A_{0,N}(z) & B_1(z) & \ldots &
    B_{\gamma-1}(z) & C_0(z) \\
\vdots & \vdots & \ddots & \vdots & \vdots & \ddots &
    \vdots & \vdots \\
\dn{N} & A_{N,0}(z) & \ldots & A_{N,N}(z) & B_1(z) & \ldots &
    B_{\gamma-1}(z) & C_N(z) \\
\up{1} & D_{1,0}(z) & \ldots & D_{1,N}(z) & 0 & \ldots & 0 & E_1(z) \\
\vdots & \vdots & \ddots & \vdots & \vdots & \ddots &
    \vdots & \vdots \\
\up{\gamma-1} & D_{\gamma-1,0}(z) & \ldots & D_{\gamma-1,N}(z) & 0 &
  \ldots & 0 & E_{\gamma-1}(z) \\
\nd & 0 & \ldots & 0 & 0 & \ldots & 0 & 0 \\
\end{block}
\end{blockarray}
\end{equation*}
where
\begin{gather}
\tag{\ref{eq:A}}
A_{m,n} =
\omega_n pz \sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i + \omega_n
pz \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \cdot
(1- \zeta_{i,m,n}) \Big) (qz)^i \\
\tag{\ref{eq:B}}
B_i(z) = \Big( \xi_i^N-\xi_{i-1}^N \Big) (qz)^i \\
\tag{\ref{eq:C}}
C_m(z) =
\sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i +
  \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \Big) (qz)^i \\
\tag{\ref{eq:D}}
D_{j,m}(z) = \omega_m pz \sum_{i=0}^{\gamma-j-1} (qz)^i \\
\tag{\ref{eq:E}}
E_j(z) = \sum_{i=0}^{\gamma-j-1} (qz)^i
\end{gather}
and where
\begin{gather}
\tag{\ref{eq:omega}}
\omega_m = {N \choose m} \big(1 - \mu/3\big)^{N-m} \big(\mu/3\big)^m \\
\tag{\ref{eq:xi}}
\xi_i = 1-(1-\mu)^i \\
\tag{\ref{eq:eta}}
\eta_i = 1-(1-\mu)^i\mu/3 \\
\tag{\ref{eq:zeta}}
\zeta_{i,m,n} = \sum_{r\geq1} {N-m \choose r}
(1-\mu)^{ri} \psi_{i,m,n,r} \bigg/ {N \choose n} \\
\notag
\psi_{i,m,n,r} = \sum_{q \geq 0}{m \choose q}{N-m-r \choose n-r-q}
\xi_i^{n-r-q}.
\end{gather}

@@

\section{Estimating the parameters}

Let us assume that we have a sequence of length $L$ that has $N$
duplicates in the genome. Let us further assume that the instrument has a
uniform error rate of $p$, and that at each nucleotide, duplicates are
different from the original sequence with probability $\mu$. What are the
values of $p$ and $\mu$ that maximize the probability that the read is
different from the original, but identical to a duplicate?

If the read has no error, this cannot happen. If we consider the reads
with one error, the probability that this happens is
\begin{equation}
Q(p,\mu) = L \cdot p \cdot (1-p)^{L-1}
\cdot \Big( 1- (1-(1-\mu)^{L-1} \mu/3)^N \Big).
\end{equation}

Observe that for every value of $p$, $\mu = 1/L$ maximizes $Q(p,\mu)$. So
in this simple proble, there exists a ``worst'' value of $\mu$, that
maximizes the probability of making a mistake irrespective of the
precision of the sequencing instrument and of the number of duplicates.

The MEM seeding heuristic is more complex than the toy problem above, but
we still get the intuition that values of $\mu$ close to $1/L$ are the
most problematic, where $L$ is the typical size of MEM seeds.

Using the above methods to compute MEM seeding probabilities, we can test
this claim with typical values of the parameters. For the human genome,
seeds of length $\approx 20$ are the norm...
